k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-03-07 paper.Rmd
abstract

    Numerous applications in psychological research require that a data set is partitioned via the maximization---instead of the more common minimization---of a clustering criterion. Such anticlustering seeks for high similarity between groups (i.e., maximum diversity) and/or high pairwise dissimilarity within groups (i.e., maximum dispersion). Brusco et al. (2020) proposed a bicriterion algorithm (BILS) that simultaneously seeks for maximum diversity and dispersion, introducing the bicriterion approach for anticlustering. Here, we advance the bicriterion approach in several ways. As a heuristic, the BILS does not necessarily identify globally optimal criterion values. We combine a novel exact algorithm for the dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. We present several adaptations of the BILS and evaluate which is best suited to accomplish this goal. Despite its theoretical computational hardness, our optimal method scales to large data sets (e.g., $N > 5000$ for $K = 3$ groups). Our analysis reveals an important factor determining the difficulty of the bicriterion anticlustering problem: the level of restriction induced by the maximum dispersion constraint. Depending on the level of restriction, different algorithmic choices are recommended. In particular, the more constrained the problem becomes, the more important is the iterated local search (ILS), the second phase of the BILS. The original BILS as well as our extensions are available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.

>>>>

    Numerous applications in psychological research require that a data set is partitioned via the maximization---instead of the more common minimization---of a clustering criterion. Such anticlustering seeks for high similarity between groups (i.e., maximum diversity) and/or high pairwise dissimilarity within groups (i.e., maximum dispersion). Brusco et al. (2020) proposed a bicriterion algorithm (BILS) that simultaneously seeks for maximum diversity and dispersion, introducing the bicriterion approach for anticlustering. Here, we advance the bicriterion approach in several ways. As our main contribution, we combine a novel exact algorithm for maximum dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. Despite its theoretical computational hardness, our optimal method scales to rather large data sets. It generates multiple partitions that the BILS can use as initializations during its local search, thus guaranteeing that the BILS also returns a partition having optimal dispersion. In a simulation study, we compared several adaptations of the BILS. To obtain high diversity, initializing the BILS with multiple optimal partitions is preferable to using only one. Moreover, in data sets where the constraint of optimal dispersion severely restricts the search for feasible partitions, the iterated local search phase of the BILS is crucial. While the original BILS oftentimes finds optimal criterion values for the dispersion, our simulation illustrates conditions when it does not. The optimal dispersion algorithm and the BILS including our extensions are available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.
